---
title: "Untitled"
author: "francesco"
date: "2024-01-09"
output:
  pdf_document: default
  html_document: default
---

Import libraries

```{r}
library(devtools)
library(footBayes)
library(bayesplot)
library(loo)
library(ggplot2)
library(dplyr)
library(tidyverse)
require(rstan)
library(ggplot2)
library(ggrepel)
library(plotly)
library(cluster)
library(knitr)
library(kableExtra)
library(dagitty)
set.seed(1)  
```





Create dataframe for multiple seasons (REMOTE data from footBayes library)
```{r}
data("italy")
italy <- as.data.frame(italy)

#italy_19_to_21 <- subset(italy[, c(2, 3, 4, 6, 7)], Season %in% c("2019", "2020", "2021"))
#colnames(italy_19_to_21) <- c("season", "home", "away", "homegoals", "awaygoals")

italy_13_to_21 <- subset(italy[, c(2, 3, 4, 6, 7)], Season %in% c("2013", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021"))
colnames(italy_13_to_21) <- c("season", "home", "away", "homegoals", "awaygoals")
head(italy_13_to_21)

```


#########################################################################################################################################################################################
#########################################################################################################################################################################################

"data_definition" function -> returns the data needed for the STAN model
```{r}

data_definition_beta <- function(data,
                      predict){
#DATA CHECK
 if (!is.matrix(data) & !is.data.frame(data)){
    stop("Data are not stored in matrix/data frame
         structure.")
  }
  
  if (dim(data)[2]<5){
    stop("Data dimensions are wrong! Supply a matrix/data frame containing the following mandatory column items:
         season, home team, away team, home goals, away goals.")
  }

  if ( !is.numeric(data$homegoals) |!is.numeric(data$awaygoals)){
    stop("Goals are not numeric!")
  }

  if (dim(data)[2]>5){
    warning("Dataset too large! The function will evaluate the first five columns as follows: season, home team, away team, home goals, away goals")
  }
  
  colnames(data) <- c("season", "home", "away", "homegoals", "awaygoals")

   
#PREDICT CHECK
  if (predict == 0){
    predict = 0
	  ngames = dim(data)[1]
	  nfit = ngames
	  npred = predict
  }else if(is.numeric(predict)){
    ngames = dim(data)[1]
    nfit = ngames-predict
    npred = predict
  }
  
  
#DATA FOR STAN MODEL
  nteams = length(unique(data$home))     #number of teams
  teams = unique(data$home)              #unique team names
  home = match(data$home, teams)         #team home (indexes of the whole ngames)
  away = match(data$away, teams)         #team away (indexes of the whole ngames)
  team1 = home[1:nfit]                   #idx home extraction for the first nfirst matches (total - predicted)
  team2 = away[1:nfit]                   #idx away extraction for the first nfirst matches (total - predicted)
  score1 = data$homegoals[1:nfit]        #score home team
  score2 = data$awaygoals[1:nfit]        #score away team
  team1pred = home[(nfit+1):(ngames)]    #home prev  || ngames = nfit + nprev
  team2pred = away[(nfit+1):(ngames)]    #away prev  || ngames = nfit + nprev
  diff_score = score1 - score2           #diff for Skellam
  

  

  data_stan = list(
    nteams = nteams,          
    teams = teams,
    home = home,
    away = away,
    team1 = team1,        
    team2 = team2,        
    team1pred = team1pred,    
    team2pred = team2pred,    
    score1 = score1,
    score2 = score2,
    diff_score = (score1-score2),
    ngames = ngames,
    nfit = nfit,
    npred = npred)
    
  return(data_stan)
}
```

Recall data_definition function providing a dataset and the number of games to predict as argument. 
```{r}
up_data_beta = data_definition_beta(italy_13_to_21, 5)
```


STAN model definition | Beta sparse (0.001, 0.001) 
```{r}
poiss_beta_expl_NI <- "
data {
  int<lower=0> nteams; //number of teams
  int<lower=0> ngames; //number of games (all)
  int<lower=0> nfit;   //number of games (for train -- [total - predict]) 
  int<lower=0> team1[nfit]; //home team index
  int<lower=0> team2[nfit]; //away team index
  int<lower=0> score1[nfit]; //score home team
  int<lower=0> score2[nfit]; //score away team
  int<lower=0> npred; //number of predicted games
  int<lower=0> team1pred[npred]; //home team index for prediction
  int<lower=0> team2pred[npred]; //away team index for prediction
}

parameters {
  vector<lower=0, upper=1>[nteams] att; //attack ability of each team
  vector<lower=0, upper=1>[nteams] def; //defence ability of each team
  real<lower=0, upper=1> home; //home advantage
  
  //hyper parameters
  //real<lower=0> alpha_att;
  //real<lower=0> alpha_def;
  //real<lower=0> beta_att;
  //real<lower=0> beta_def;
}

transformed parameters {
  //vector[nteams] att; 
  //vector[nteams] def; 
  vector[nfit] theta1; //score intensity of home team   | test con vector[nfit] per non ottenere elementi vuoti
  vector[nfit] theta2; //score intensity of away team   | test con vector[nfit] per non ottenere elementi vuoti

  //for (t in 1:nteams){
    //att[t] = att_raw[t];  //-mean(att_raw);
    //def[t] = def_raw[t];  //-mean(def_raw);
  //}

  theta1 = exp(home + att[team1] - def[team2]);
  theta2 = exp(att[team2] - def[team1]);
}

model {
  // priors
  for (t in 1:nteams) {
    att[t] ~ beta(0.001, 0.001);
    def[t] ~ beta(0.001, 0.001);
  }

  home ~ beta(1, 1);  

  // hyper-priors
  //alpha_att ~ beta(0.1, 0.1);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  //alpha_def ~ beta(0.1, 0.1);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  //beta_att ~ beta(0.1, 0.1);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  //beta_def ~ beta(0.1, 0.1);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  // likelihood
  score1 ~ poisson(theta1);
  score2 ~ poisson(theta2);
}


generated quantities {
//generate predictions
  //real s1rep[nfit]; //replicated score //in-sample
  //real s2rep[nfit]; //replicated score //in-sample
  real s1pred[npred]; //predicted score //o-o-sample
  real s2pred[npred]; //predicted score //o-o-sample
  vector[npred] theta1pred; //score probability of home team //o-o-sample
  vector[npred] theta2pred; //score probability of away team //o-o-sample

  //s1rep = poisson_rng(theta1);  //in-sample
  //s2rep = poisson_rng(theta2);  //in-sample

  theta1pred = exp(home + att[team1pred] - def[team2pred]); //o-o-sample
  theta2pred = exp(att[team2pred] - def[team1pred]); //o-o-sample
   
  s1pred = poisson_rng(theta1pred);  //o-o-sample
  s2pred = poisson_rng(theta2pred);  //o-o-sample
  
}
"

```

STAN model definition | Beta (1, 1)
```{r}
poiss_beta_expl <- "
data {
  int<lower=0> nteams; //number of teams
  int<lower=0> ngames; //number of games (all)
  int<lower=0> nfit;   //number of games (for train -- [total - predict]) 
  int<lower=0> team1[nfit]; //home team index
  int<lower=0> team2[nfit]; //away team index
  int<lower=0> score1[nfit]; //score home team
  int<lower=0> score2[nfit]; //score away team
  int<lower=0> npred; //number of predicted games
  int<lower=0> team1pred[npred]; //home team index for prediction
  int<lower=0> team2pred[npred]; //away team index for prediction
}

parameters {
  vector<lower=0, upper=1>[nteams] att; //attack ability of each team
  vector<lower=0, upper=1>[nteams] def; //defence ability of each team
  real<lower=0, upper=1> home; //home advantage
  
  //hyper parameters
  //real<lower=0> alpha_att;
  //real<lower=0> alpha_def;
  //real<lower=0> beta_att;
  //real<lower=0> beta_def;
}

transformed parameters {
  //vector[nteams] att; 
  //vector[nteams] def; 
  vector[nfit] theta1; //score intensity of home team   | test con vector[nfit] per non ottenere elementi vuoti
  vector[nfit] theta2; //score intensity of away team   | test con vector[nfit] per non ottenere elementi vuoti

  //for (t in 1:nteams){
    //att[t] = att_raw[t];  //-mean(att_raw);
    //def[t] = def_raw[t];  //-mean(def_raw);
  //}

  theta1 = exp(home + att[team1] - def[team2]);
  theta2 = exp(att[team2] - def[team1]);
}

model {
  // priors
  for (t in 1:nteams) {
    att[t] ~ beta(1, 1);
    def[t] ~ beta(1, 1);
  }

  home ~ beta(1, 1);  

  // hyper-priors
  //alpha_att ~ beta(0.1, 0.1);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  //alpha_def ~ beta(0.1, 0.1);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  //beta_att ~ beta(0.1, 0.1);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  //beta_def ~ beta(0.1, 0.1);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  // likelihood
  score1 ~ poisson(theta1);
  score2 ~ poisson(theta2);
}


generated quantities {
//generate predictions
  //real s1rep[nfit]; //replicated score //in-sample
  //real s2rep[nfit]; //replicated score //in-sample
  real s1pred[npred]; //predicted score //o-o-sample
  real s2pred[npred]; //predicted score //o-o-sample
  vector[npred] theta1pred; //score probability of home team //o-o-sample
  vector[npred] theta2pred; //score probability of away team //o-o-sample

  //s1rep = poisson_rng(theta1);  //in-sample
  //s2rep = poisson_rng(theta2);  //in-sample

  theta1pred = exp(home + att[team1pred] - def[team2pred]); //o-o-sample
  theta2pred = exp(att[team2pred] - def[team1pred]); //o-o-sample
   
  s1pred = poisson_rng(theta1pred);  //o-o-sample
  s2pred = poisson_rng(theta2pred);  //o-o-sample
  
}
"
```

STAN model definition | Beta informative ( 5, 5)
```{r}
poiss_beta_expl_I <- "
data {
  int<lower=0> nteams; //number of teams
  int<lower=0> ngames; //number of games (all)
  int<lower=0> nfit;   //number of games (for train -- [total - predict]) 
  int<lower=0> team1[nfit]; //home team index
  int<lower=0> team2[nfit]; //away team index
  int<lower=0> score1[nfit]; //score home team
  int<lower=0> score2[nfit]; //score away team
  int<lower=0> npred; //number of predicted games
  int<lower=0> team1pred[npred]; //home team index for prediction
  int<lower=0> team2pred[npred]; //away team index for prediction
}

parameters {
  vector<lower=0, upper=1>[nteams] att; //attack ability of each team
  vector<lower=0, upper=1>[nteams] def; //defence ability of each team
  real<lower=0, upper=1> home; //home advantage
  
  //hyper parameters
  //real<lower=0> alpha_att;
  //real<lower=0> alpha_def;
  //real<lower=0> beta_att;
  //real<lower=0> beta_def;
}

transformed parameters {
  //vector[nteams] att; 
  //vector[nteams] def; 
  vector[nfit] theta1; //score intensity of home team   | test con vector[nfit] per non ottenere elementi vuoti
  vector[nfit] theta2; //score intensity of away team   | test con vector[nfit] per non ottenere elementi vuoti

  //for (t in 1:nteams){
    //att[t] = att_raw[t];  //-mean(att_raw);
    //def[t] = def_raw[t];  //-mean(def_raw);
  //}

  theta1 = exp(home + att[team1] - def[team2]);
  theta2 = exp(att[team2] - def[team1]);
}

model {
  // priors
  for (t in 1:nteams) {
    att[t] ~ beta(5, 5);
    def[t] ~ beta(5, 5);
  }

  home ~ beta(1, 1);  

  // hyper-priors
  //alpha_att ~ beta(0.1, 0.1);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  //alpha_def ~ beta(0.1, 0.1);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  //beta_att ~ beta(0.1, 0.1);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  //beta_def ~ beta(0.1, 0.1);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  // likelihood
  score1 ~ poisson(theta1);
  score2 ~ poisson(theta2);
}


generated quantities {
//generate predictions
  //real s1rep[nfit]; //replicated score //in-sample
  //real s2rep[nfit]; //replicated score //in-sample
  real s1pred[npred]; //predicted score //o-o-sample
  real s2pred[npred]; //predicted score //o-o-sample
  vector[npred] theta1pred; //score probability of home team //o-o-sample
  vector[npred] theta2pred; //score probability of away team //o-o-sample

  //s1rep = poisson_rng(theta1);  //in-sample
  //s2rep = poisson_rng(theta2);  //in-sample

  theta1pred = exp(home + att[team1pred] - def[team2pred]); //o-o-sample
  theta2pred = exp(att[team2pred] - def[team1pred]); //o-o-sample
   
  s1pred = poisson_rng(theta1pred);  //o-o-sample
  s2pred = poisson_rng(theta2pred);  //o-o-sample
  
}
"
```

STAN model definition | Beta Beta (0.001, 0.001) 
```{r}
poiss_beta_beta <- "
data {
  int<lower=0> nteams; //number of teams
  int<lower=0> ngames; //number of games (all)
  int<lower=0> nfit;   //number of games (for train -- [total - predict]) 
  int<lower=0> team1[nfit]; //home team index
  int<lower=0> team2[nfit]; //away team index
  int<lower=0> score1[nfit]; //score home team
  int<lower=0> score2[nfit]; //score away team
  int<lower=0> npred; //number of predicted games
  int<lower=0> team1pred[npred]; //home team index for prediction
  int<lower=0> team2pred[npred]; //away team index for prediction
}

parameters {
  vector<lower=0, upper=1>[nteams] att; //attack ability of each team
  vector<lower=0, upper=1>[nteams] def; //defence ability of each team
  real<lower=0, upper=1> home; //home advantage
  
  //hyper parameters
  real<lower=0, upper=1> alpha_att;
  real<lower=0, upper=1> alpha_def;
  real<lower=0, upper=1> beta_att;
  real<lower=0, upper=1> beta_def;
}

transformed parameters {
  //vector[nteams] att; 
  //vector[nteams] def; 
  vector[nfit] theta1; //score intensity of home team   | test con vector[nfit] per non ottenere elementi vuoti
  vector[nfit] theta2; //score intensity of away team   | test con vector[nfit] per non ottenere elementi vuoti

  theta1 = exp(home + att[team1] - def[team2]);
  theta2 = exp(att[team2] - def[team1]);
}

model {
  // priors
  for (t in 1:nteams) {
    att[t] ~ beta(alpha_att, beta_att);
    def[t] ~ beta(alpha_def, beta_def);
  }

  home ~ beta(1, 1);  

// hyper-priors
  alpha_att ~ beta(0.001, 0.001);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  alpha_def ~ beta(0.001, 0.001);   // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  beta_att ~ beta(0.001, 0.001);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE
  beta_def ~ beta(0.001, 0.001);    // (0.5, 0.5) | (1, 1) NON-INFORMATIVE |(2, 2) (10, 10) #INFORMATIVE

  // likelihood
  score1 ~ poisson(theta1);
  score2 ~ poisson(theta2);
}


generated quantities {
//generate predictions
  //real s1rep[nfit]; //replicated score //in-sample
  //real s2rep[nfit]; //replicated score //in-sample
  real s1pred[npred]; //predicted score //o-o-sample
  real s2pred[npred]; //predicted score //o-o-sample
  vector[npred] theta1pred; //score probability of home team //o-o-sample
  vector[npred] theta2pred; //score probability of away team //o-o-sample

  //s1rep = poisson_rng(theta1);  //in-sample
  //s2rep = poisson_rng(theta2);  //in-sample

  theta1pred = exp(home + att[team1pred] - def[team2pred]); //o-o-sample
  theta2pred = exp(att[team2pred] - def[team1pred]); //o-o-sample
   
  s1pred = poisson_rng(theta1pred);  //o-o-sample
  s2pred = poisson_rng(theta2pred);  //o-o-sample
  
}
"

```



Run of the model
```{r}
writeLines(poiss_beta_expl_I, "poiss_beta_expl_I.stan")
first_model_beta = stan(file ="poiss_beta_expl_I.stan",  data = up_data_beta, verbose = FALSE)
```


Model parameters extraction
```{r}
model_param_beta = rstan::extract(first_model_beta)
```


```{r}
plot(model_param$home)
plot(model_param_beta$home)

# Crea un dataframe con i dati
data_boxplot <- data.frame(
  Method = rep(c("GP model", "Beta model"), each = length(model_param$home)),
  Values = c(model_param$home, model_param_beta$home)
)


ggplot(data_boxplot, aes(x = Method, y = Values, fill = Method)) +
  geom_boxplot() +
  labs(title = "Boxplot Home parameter",
       x = "Models",
       y = "Values") +
  scale_fill_manual(values = c("GP model" = "lightgreen", "Beta model" = "orange"))

  
ggsave("C:/Users/kecco/Desktop/TESI_magistrale/images_thesis/boxplot_home_twomodels.png", width = 8, height = 6, dpi = 300)


```



```{r}
library(ggplot2)

# Assuming plot_data is your vector of data
home_data <- model_param_beta$home

# Create the boxplot
boxplot <- ggplot(data.frame(value = home_data), aes(y = value)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Boxplot of Home Parameter (Beta(1, 1))", y = "Home Parameter") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white"),  # Set background color
    plot.background = element_rect(fill = "white"),   # Set plot background color
    panel.grid.major = element_line(color = "gray", linetype = "dashed"),  # Add grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.line = element_line(color = "black"),  # Set axis line color
    text = element_text(color = "black")  # Set text color
  )

# Save the boxplot
ggsave("C:/Users/kecco/Desktop/TESI_magistrale/images_thesis/home_boxplot_Beta_1_1.png", boxplot, width = 8, height = 6, dpi = 300)


```



Plot for predicted games
```{r}
pred_scores = c(colMeans(model_param$s1pred),colMeans(model_param$s2pred)) 
#true_scores = c(italy_13_to_21$homegoals[(up_data$nfit+1):up_data$ngames],italy_13_to_21$awaygoals[(up_data$nfit+1):up_data$ngames])
true_scores = c(italy_13_to_21$homegoals[(up_data$nfit+1):up_data$ngames],italy_13_to_21$awaygoals[(up_data$nfit+1):up_data$ngames])
plot(true_scores, pred_scores, xlim=c(0,5), ylim=c(0,5), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')

pred_errors = c(sapply(1:up_data$npred, function(x) sd(model_param$s1pred[,x])),sapply(1:up_data$npred, function(x) sd(model_param$s2pred[,x])))
arrows(true_scores, pred_scores+pred_errors, true_scores, pred_scores-pred_errors, length = 0.05, angle = 90, code = 3, col=rgb(0,0,0,0.3))
```


Season target definition -> returns the names of the teams participating in Serie A in the season of interest.
```{r}
#unique_ss <- unique(subset(italy_13_to_21$home, italy_13_to_21$season == "2021"))
unique_ss <- unique(subset(italy_13_to_21$home, italy_13_to_21$season == "2021"))

```


Plot for the att - def capability of the teams

```{r}
# attacchi e le difese come media e deviazione standard
attack <- colMeans(model_param_beta$att)
attacksd <- sapply(1:up_data_beta$nteams, function(x) sd(model_param_beta$att[, x]))
defense <- colMeans(model_param_beta$def)
defensesd <- sapply(1:up_data_beta$nteams, function(x) sd(model_param_beta$def[, x]))

# filtro gli attacchi e le difese solo per le squadre in unique_ss
filtered_attack <- attack[up_data_beta$teams %in% unique_ss]
filtered_defense <- defense[up_data_beta$teams %in% unique_ss]
filtered_attacksd <- attacksd[up_data_beta$teams %in% unique_ss]
filtered_defensesd <- defensesd[up_data_beta$teams %in% unique_ss]

data <- data.frame(
  Attack = filtered_attack,
  Defense = filtered_defense,
  Team = up_data_beta$teams[up_data_beta$teams %in% unique_ss]
)

```

```{r}
# ggplot2 e ggrepel
library(ggplot2)
library(ggrepel)

# Create a data frame with mean and standard deviation values
data <- data.frame(
  Attack = filtered_attack,
  Defense = filtered_defense,
  attacksd = filtered_attacksd,
  defensesd = filtered_defensesd,
  Team = up_data_beta$teams[up_data_beta$teams %in% unique_ss]
)
```



```{r}
k <- 4

# Perform K-means clustering
set.seed(123)  # Set seed for reproducibility
data$cluster <- kmeans(data[, c("Defense", "Attack")], centers = k)$cluster

# Create a ggplot scatter plot
p <- ggplot(data, aes(x = Defense, y = Attack, label = Team, color = factor(cluster))) +
  labs(x = "Defense", y = "Attack", title = "Attack & Defense") +
  
  # Add arrows for standard deviations
  geom_errorbar(aes(x = Defense, ymin = Attack - attacksd, ymax = Attack + attacksd), width = 0, linetype = "dashed",  color = "darkgrey", alpha = 0.5) +
  geom_errorbarh(aes(y = Attack, xmin = Defense - defensesd, xmax = Defense + defensesd), height = 0, linetype = "dashed", color = "darkgrey", alpha = 0.5) +
  
  # Add labels with repel to avoid overlapping
  geom_text_repel(aes(color = factor(cluster)), box.padding = 0.63, size = 3, max.overlaps = Inf, color = "black") +  # Adjust 'size' and 'max.overlaps' as needed
  geom_point(size = 3.5, alpha = 0.7) +  # Adjust 'size' and 'alpha' as needed

  # Set theme with a white background
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  ) +
  
  # Add cluster colors
  scale_color_manual(name = "Cluster", values = c("red", "blue", "green", "orange")) +
  
  # Add dashed line at x = 0.5 and y = 0.5
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black") +
  
  # Set X and Y axis limits
  xlim(0, 1) +
  ylim(0, 1)



print(p)

ggsave("C:/Users/kecco/Desktop/TESI_magistrale/images_thesis/Beta_final_model_2021.png", width = 8, height = 6, dpi = 300)

```













Testing different distances between coordinates of each point in order to have an overall ranking  

```{r}
# Euclidean: radice quadrata della somma dei quadrati delle differenze tra le coordinate dei punti
euclidean_distance <- sqrt(filtered_attack^2 + filtered_defense^2)

# Manhattan: somma delle differenze assolute tra le coordinate dei punti
manhattan_distance <- abs(filtered_attack - 0) + abs(filtered_defense - 0)

```

```{r}
# Creazione del dataframe
distance_euc <- data.frame(
  Team = up_data$teams[up_data$teams %in% unique_ss],
  Euclidean = euclidean_distance
)

# Creazione del dataframe
distance_man <- data.frame(
  Team = up_data$teams[up_data$teams %in% unique_ss],
  Manhattan = manhattan_distance
)


# Ordina il dataframe in base alla colonna Euclidean in ordine decrescente
distance_euc <- arrange(distance_euc, desc(Euclidean))
distance_man <- arrange(distance_man, desc(Manhattan))


# Stampa del dataframe ordinato
print(distance_euc)
print(distance_man)

```

Viewing in the same dataframe
```{r}
# Combinazione dei dataframe mantenendo solo le colonne necessarie
distance_combined <- merge(distance_euc, distance_man, by = "Team")[c("Team", "Euclidean", "Manhattan")]

# Ordina il dataframe combinato in base alla colonna Euclidean in ordine decrescente
distance_combined <- arrange(distance_combined, desc(Euclidean))

# Stampa del dataframe combinato ordinato con stile
show(distance_combined)
```
